---
Title: "A Guide on Trajectory Analysis"
Author: "Rufai Omowunmi Balogun, Sunita Phuyal, Daniel Neau, Harry Leduc"
Description: This is the tutorial for the trajectory analysis of the area of interest in R from the beginning  
Date: "2024-11-15"
Output: html_document
---


```{r, include= FALSE}
library(terra)
library(sp)
library(raster)
```

## We need to give the folder path to the data. Here, we are using the Mahakam Delta, Indonesia, where the mangroves are converted into shrimp pond. 

```{r, include=TRUE}
raster_mahakam1999 <- rast("C:/Users/s_phu/Documents/TrajectoryAnalysisTutorial/Data/Mahakam_1999.rst")
raster_mahakam2014 <- rast("C:/Users/s_phu/Documents/TrajectoryAnalysisTutorial/Data/Mahakam_2014.rst")
raster_mahakam2018 <- rast("C:/Users/s_phu/Documents/TrajectoryAnalysisTutorial/Data/Mahakam_2018.rst")
raster_mahakam2020 <- rast("C:/Users/s_phu/Documents/TrajectoryAnalysisTutorial/Data/Mahakam_2020.rst")
raster_mahakam2022 <- rast("C:/Users/s_phu/Documents/TrajectoryAnalysisTutorial/Data/Mahakam_2022.rst")

# Extract raster values, count the frequency 
raster_1999 <- values(raster_mahakam1999)
frequency_1999 <- table(raster_1999)
print(frequency_1999)

raster_2014 <- values(raster_mahakam2014)
frequency_2014 <- table(raster_2014)
print(frequency_2014)

raster_2018 <- values(raster_mahakam2018)
frequency_2018 <- table(raster_2018)
print(frequency_2018)

raster_2020 <- values(raster_mahakam2020)
frequency_2020 <- table(raster_2020)
print(frequency_2020)

raster_2022 <- values(raster_mahakam2022)
frequency_2022 <- table(raster_2022)
print(frequency_2022)
```

## Now we can reclassify the raster file as per our objectives. 
## Here, we are creating boolean map

```{r, include=TRUE}
## For 1999
crs(raster_mahakam1999) <- "EPSG:32749"
raster_1999 <- raster_mahakam1999
raster_1999[!is.na(raster_1999) & raster_1999 == 1] <- 1  # Mangrove
raster_1999[!is.na(raster_1999) & raster_1999 != 1] <- 0  # Other
raster_1999_values <- values(raster_mahakam1999)
raster_1999_values[!is.na(raster_1999_values) & raster_1999_values == 1] <- 1
raster_1999_values[!is.na(raster_1999_values) & raster_1999_values != 1] <- 0
mangrove_1999 <- setValues(raster_mahakam1999, raster_1999_values)

##  For 2014
crs(raster_mahakam2014) <- "EPSG:32749"
raster_2014 <- raster_mahakam2014
raster_2014[!is.na(raster_2014) & raster_2014 == 1] <- 1  # Mangrove
raster_2014[!is.na(raster_2014) & raster_2014 != 1] <- 0  # Other
raster_2014_values <- values(raster_mahakam2014)
raster_2014_values[!is.na(raster_2014_values) & raster_2014_values == 1] <- 1
raster_2014_values[!is.na(raster_2014_values) & raster_2014_values != 1] <- 0
mangrove_2014 <- setValues(raster_mahakam2014, raster_2014_values)

# For 2018
crs(raster_mahakam2018) <- "EPSG:32749"
raster_2018 <- raster_mahakam2018
raster_2018[!is.na(raster_2018) & raster_2018 == 1] <- 1  # Mangrove
raster_2018[!is.na(raster_2018) & raster_2018 != 1] <- 0  # Other
raster_2018_values <- values(raster_mahakam2018)
raster_2018_values[!is.na(raster_2018_values) & raster_2018_values == 1] <- 1
raster_2018_values[!is.na(raster_2018_values) & raster_2018_values != 1] <- 0
mangrove_2018 <- setValues(raster_mahakam2018, raster_2018_values)

## For 2020
crs(raster_mahakam2020) <- "EPSG:32749"
raster_2020 <- raster_mahakam2020
raster_2020[!is.na(raster_2020) & raster_2020 == 1] <- 1  # Mangrove
raster_2020[!is.na(raster_2020) & raster_2020 != 1] <- 0  # Other
raster_2020_values <- values(raster_mahakam2020)
raster_2020_values[!is.na(raster_2020_values) & raster_2020_values == 1] <- 1
raster_2020_values[!is.na(raster_2020_values) & raster_2020_values != 1] <- 0
mangrove_2020 <- setValues(raster_mahakam2020, raster_2020_values)

## For 2022
crs(raster_mahakam2022) <- "EPSG:32749"
raster_2022 <- raster_mahakam2022
raster_2022[!is.na(raster_2022) & raster_2022 == 1] <- 1  # Mangrove
raster_2022[!is.na(raster_2022) & raster_2022 != 1] <- 0  # Other
raster_2022_values <- values(raster_mahakam2022)
raster_2022_values[!is.na(raster_2022_values) & raster_2022_values == 1] <- 1
raster_2022_values[!is.na(raster_2022_values) & raster_2022_values != 1] <- 0
mangrove_2022 <- setValues(raster_mahakam2022, raster_2022_values)
```

## Creating stack map 
```{r, include=TRUE}
raster_stack <- c(mangrove_1999, mangrove_2014, mangrove_2018, mangrove_2020, mangrove_2022)
names(raster_stack) <- c ("1999", "2014", "2018", "2020", "2022")

color_palette <- c("blue", "red")

plot(raster_stack,
     col = color_palette,
     main = "Mangrove Trajectory (1999 - 2022)",
     axes = TRUE)
writeRaster(raster_stack, "Mangrove_Trajectory_1999_2022.tif", overwrite = TRUE)

## Since we have Created the Boolean Map, we will be working on these maps for Trajectory Analysis
rasstackY <- terra::rast("Mangrove_Trajectory_1999_2022.tif")
rasstackY

plot(rasstackY)
```

# Let's dive into the Trajectory Analysis


```{r, include=FALSE}
##lets load the timeseriesTrajectories package

library(timeseriesTrajectories)
```

```{r, include=TRUE}
## reclassifying so that 0 becomes 1 and 1 becomes 2
reclass_df <- cbind(from = c(1), 
                    to = c(2), 
                    becomes = c(2))
rasstackY_v2 <- classify(rasstackY,reclass_df,include.lowest = TRUE)
plot(rasstackY_v2)

##Creating time series plot from dataPreview function
tps = c(1999,2014,2018,2020,2022)

# I know the units of my variable is number of pixels, thus I will create a sting for the vertical units.
vert_units <- "number of pixels"

# Next, pass the variables to the dataPreview function.
# NB: xAngle sets the orientation of the horizontal axis labels. See help file for more details.
dataPreview(rasstackY,
            timepoints = tps,
            vertunits = vert_units,
            xAngle = 0)

## Creating data for the number of presence and change
num_pres_change <- presenceData(rasstackY, nodata = -999)
num_pres_change

## Creating map for the number of presence and change
presencePlot(num_pres_change,
             pltunit = "m",
             dataEpsg = 32749,
             scalePos = "bottomright",
             narrowPos = "topleft",
             narrowSize = 1,
             categoryName = "Mangrove",
             xAxis = "Longitude (m)",
             yAxis = "Latitude (m)",
             axisText = 1,
             axisLabel = 1,
             plotTitle = 1.2)

## Creating input data for trajectory plot
traj_data <- rastertrajData(rasstackY,
                            zeroabsence = 'yes')
traj_data

## Creating a map and chart of the trajectory of Mahakam Delta
trajPlot(traj_data,
         axisShow = "no",
         categoryName = "Mangrove",
         narrowPos = NA,
         scalePos = NA,
         scaleSize = 1.5,
         axisText = 1.2,
         axisLabel = 1.2,
         plotTitle = 1,
         legendTex = 1,
         xAxis = "Longitude (m)",
         yAxis = "Latitude (m)",
         downsample = TRUE)

## Creating input data for plot
tps = c(2000,2001,2002,2003,2005) ##converting into vector for the time point
stackbar_data <- rasterstackData(rasstackY,
                                 timePoints = c(1999,2014,2018,2020,2022),
                                 spatialextent = 'unified',
                                 zeroabsence = 'yes',
                                 annualchange = 'yes',
                                 categoryName = 'variable',
                                 regionName = 'region',
                                 varUnits = "(squre kilometers)",
                                 constant  = 1)
stackbar_data

##Now, plotting these data
stackbarPlot(stackbar_data,
             axisSize = 10,
             lbAxSize = 10,
             lgSize = 7.5,
             titleSize = 12,
             datbreaks = "no",
             upperlym = 35,
             lowerlym = - 50,
             lymby = 5,
             upperlym2 = 0.5,
             lymby2 = 0.1,
             xAngle = 0)

```

##In this way we created trajectory of the mangrove.
## Reference: Bilintoh, T.M., Pontius Jr, R,.G., & Zhang, A., (2024). Methods to compare sites concerning a Category’s change during various time intervals. GIScience & Remote Sensing, 61 (1), 14. https://doi.org/10.1080/15481603.2024.2409484